---
title: "Week 4 · Class 1: Practice Exercises"
subtitle: "Transforming Variables & Causality"
date: "2026"
format:
  pdf:
    toc: false
    number-sections: true
    colorlinks: true
---

```{r}
#| include: false
library(tidyverse)
library(knitr)
library(formatR)
# When running chunks interactively, set the working directory to this file's
# folder so the relative data/ paths resolve. (During render, knitr already
# uses the document's directory, so this is skipped.)
if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
  .doc_path <- rstudioapi::getSourceEditorContext()$path
  if (length(.doc_path) == 1 && nzchar(.doc_path)) setwd(dirname(.doc_path))
}
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = FALSE)
```

# Segment A — Transforming and Creating Variables
## Worked Example: Recoding with case_when() (read this first)

Before the exercises, here is one complete transformation, start to finish. Use it as your template.

**Goal**: turn a numeric income variable (in dollars) into a three-level bracket, then count respondents per bracket.

```r
survey <- survey %>%
  mutate(
    income_k = income / 1000,                  # step 1: put income in $1000s
    bracket = case_when(                       # step 2: three+ outcomes -> case_when()
      income_k < 40              ~ "Lower",
      income_k >= 40 & income_k < 100 ~ "Middle",
      income_k >= 100            ~ "Upper"
    )
  )

survey %>% count(bracket)                      # step 3: always check your new variable
```

**Why each piece matters:**

1. `mutate()` adds columns and keeps every row — the data does not shrink.
2. Each `case_when()` line is `condition ~ label`. Conditions are checked **top to bottom**; the first TRUE wins.
3. Every condition must evaluate to TRUE or FALSE. A row matching *no* condition becomes `NA` — that is a signal to check your logic, not an error.
4. The final `count()` is the habit that catches mistakes: if a bracket has zero rows or `NA` appears, a condition is wrong.

The exercises below follow this same shape. When you use AI for the later ones, paste this example into your prompt as the style you want.

# Non-AI Exercises

## 1. Understanding mutate()

### 1.1 Multiple Choice: mutate() Function

What does the `mutate()` function do?

a) Removes columns from a data frame
b) Creates new columns or modifies existing ones
c) Filters rows based on conditions
d) Sorts data by a variable

Answer: __b___

### 1.2 Code Detective: Basic mutate()

What does this code create?

```r
data %>%
  mutate(
    vote_margin = dem_votes - rep_votes,
    winner = if_else(vote_margin > 0, "Democrat", "Republican")
  )
```

Line 3 creates: a column called "vote_margin" that takes the difference between the number of democrat votes and republican votes
Line 4 creates: a column that displays the winning party by determining if the value of the previous value is positive or not

### 1.3 Fill in the Blanks: Variable Types

When creating new variables, we often need to:

1. Calculate differences between existing variables
2. Recode text variables into categories
3. Create logical (TRUE/FALSE) indicators
4. Convert between data types
5. Handle missing values appropriately

Word bank: differences, recode, logical, data, missing

## 2. Conditional Logic

### 2.1 Match: if_else() vs case_when()

Match each function with when to use it:

**Functions:**
a) if_else()
b) case_when()

**Use cases:**
1. Creating a variable with only two possible outcomes
2. Creating a variable with multiple categories
3. Simple yes/no binary coding
4. Complex multi-condition logic

Matches: a = __1___ and ___3__, b = __2___ and __4___

### 2.2 Code Detective: case_when()

What categories does this code create?

```r
data %>%
  mutate(
    age_group = case_when(
      age < 30 ~ "Young",
      age < 50 ~ "Middle",
      age < 65 ~ "Older",
      TRUE ~ "Senior"
    )
  )
```

For someone age 25: Young
For someone age 45: Middle
For someone age 70: Senior

### 2.3 Spot the Error

What's wrong with this case_when() statement?

```r
mutate(
  income_cat = case_when(
    income < 30000 ~ "Low",
    income < 50000 ~ "Medium",
    income < 30000 ~ "Low",
    TRUE ~ "High"
  )
)
```

Problem: Line 146 is not going to produce any value and therefore redundent

## 3. Working with Proportions

### 3.1 Multiple Choice: Calculating Proportions

To calculate the proportion of Democrats in a dataset, you would:

a) Count Democrats and divide by Republicans
b) Count Democrats and divide by total observations
c) Count total and divide by Democrats
d) Use mean() on a logical variable

Answer: __B___

### 3.2 True or False: summarise()

Mark each statement as True (T) or False (F):

___T__ summarise() reduces multiple rows to a single summary row
___T__ You can calculate multiple statistics in one summarise() call
___T__ summarise() automatically groups by all variables
___T__ n() counts the number of rows in each group
___F__ summarise() can only calculate numeric summaries

### 3.3 Fill in the Code

Complete this code to calculate turnout rate:

```r
data %>%
  summarise(
    total_voters = n(),
    total_voted = sum(voted == "Yes"),
    turnout_rate = toal_voted / total_voters
  )
```

## 4. Advanced Transformations

### 4.1 Match: Functions and Purposes

Match each function with its purpose:

**Functions:**
a) round()
b) log()
c) sqrt()
d) abs()
e) lag()

**Purposes:**
1. Remove decimal places
2. Transform skewed data
3. Calculate square root
4. Make negative values positive
5. Get previous row's value

Matches: a = __1__, b = __2_, c = __3___, d = ___4__, e = __5___

### 4.2 Code Detective: Complex Transformation

What does this code calculate?

```r
data %>%
  group_by(state) %>%
  mutate(
    state_avg = mean(income, na.rm = TRUE),
    income_diff = income - state_avg,
    above_avg = income_diff > 0
  )
```

This code calculates: For each state, it calculates the average income, then for each person it calculates how their income differs from their state’s average, and creates a logical variable indicating if they earn above their state’s average

# AI Exercises

**Tips for Working with Claude:**

- Ask for **R code using only tidyverse** (no other packages)
- Request **simple, focused answers** to your specific question—not complex analyses
- Ask Claude to **explain what the code is doing** since you're learning
- Avoid asking for visualizations or plots in these exercises
- Include the output of `glimpse()` in your prompt so Claude knows your variable names

**Example prompt:** "Using tidyverse in R, create a new variable called ideology_category using case_when() that categorizes ideology_score into 'left', 'center', and 'right' based on terciles. Keep the code simple and explain what each line does. Here is what my data looks like: [paste glimpse output]"

For each AI exercise:
- Work with Claude to analyze the data
- Record your prompts and key findings

## 5. Creating Political Variables

**Dataset: data/legislative_votes.csv**

**Description**: Roll call voting data with legislator information.

**Variables**:
- `legislator_id`: Unique identifier (int)
- `name`: Legislator name (chr)
- `party`: Political party (chr)
- `state`: State abbreviation (chr)
- `district`: District number (int)
- `ideology_score`: -1 (liberal) to 1 (conservative) (dbl)
- `vote_attendance`: Percentage of votes attended (dbl)
- `bills_sponsored`: Number of bills sponsored (int)
- `years_service`: Years in office (int)
- `committee_count`: Number of committees served on (int)

### 5.1 Initial Data Exploration

```{r}
# Load the dataset
legislators <- read_csv("data/legislative_votes.csv")

# Examine the data
glimpse(legislators)
```

### 5.2 Creating Categorical Variables

Ask Claude to help you create:
- An "ideology_category" variable (left, center, right) based on terciles of `ideology_score`
- An "effectiveness" score based on bills sponsored and committee work (assume more than 4 is effective)
- A "seniority" category based on years of service (>20 is senior)

```{r}
legislators <- legislators %>%
mutate(
ideology_category = case_when(
ideology_score < quantile(ideology_score, 1/3, na.rm = TRUE) ~ "left",
ideology_score < quantile(ideology_score, 2/3, na.rm = TRUE) ~ "center",
TRUE ~ "right"
),
effectiveness = if_else(bills_sponsored > 4 & committee_count > 0,
"effective", "not effective"),
seniority = if_else(years_service > 20, "senior", "junior")
)

legislators %>%
count(ideology_category)
```

### 5.3 Calculating Party Metrics

Work with Claude to calculate party-level statistics for the variables you created in the last step.  For each party, what percent of legislators are in each category you created.

```{r}
party_stats <- legislators %>%
group_by(party) %>%
summarise(
# Ideology distribution
pct_left = mean(ideology_category == "left", na.rm = TRUE) * 100,
pct_center = mean(ideology_category == "center", na.rm = TRUE) * 100,
pct_right = mean(ideology_category == "right", na.rm = TRUE) * 100,
# Effectiveness
pct_effective = mean(effectiveness == "effective", na.rm = TRUE) * 100,
# Seniority
pct_senior = mean(seniority == "senior", na.rm = TRUE) * 100,
n = n()
) %>%
mutate(across(where(is.numeric) & !n, ~round(., 1)))
party_stats
```

## 6. Education and Political Participation

**Dataset: data/civic_engagement.csv**

**Description**: Survey data on education and political participation.

**Variables**:
- `respondent_id`: Unique identifier (int)
- `education`: Highest degree earned (chr)
- `age`: Age in years (int)
- `income`: Annual income (dbl)
- `voted_2020`: Whether voted in 2020 (chr: Yes/No)
- `political_interest`: 1-10 scale (int)
- `volunteer_hours`: Political volunteer hours per month (int)
- `donations`: Political donations in dollars (dbl)
- `social_media_political`: Hours per week on political social media (dbl)

### 6.1 Loading and Initial Transformation

```{r}
# Load the dataset
civic <- read_csv("data/civic_engagement.csv")

# Check the structure
glimpse(civic)
```

### 6.2 Education and Engagement Index

Work with Claude to:
- Recode education into "High School + " and "Less than High School"
- Create a variable called `activist` that is TRUE if the person donated more than the median, volunteered more than the median and voted in 2020.

```{r}
median_donations <- median(civic$donations, na.rm = TRUE)
median_volunteer <- median(civic$volunteer_hours, na.rm = TRUE)
# Create new variables
civic <- civic %>%
mutate(
# Recode education into binary
education_binary = if_else(
education %in% c("High School", "Some College", "Bachelor", "Graduate"),
"High School +",
"Less than High School"
),
# Create activist variable
# TRUE if donated > median AND volunteered > median AND voted in 2020
activist = (donations > median_donations) &
(volunteer_hours > median_volunteer) &
(voted_2020 == "Yes")
)
# Check the results
civic %>%
count(education_binary)
```

## 7. Campaign Finance Transformations

**Dataset: data/campaign_finance_2024.csv**

**Description**: Campaign contribution data for 2024 elections.

**Variables**:
- `contribution_id`: Unique identifier (int)
- `candidate`: Candidate name (chr)
- `party`: Political party (chr)
- `office`: Office sought (chr)
- `amount`: Contribution amount (dbl)
- `date`: Date of contribution (date)
- `contributor_type`: Individual, PAC, etc. (chr)
- `state`: Contributor state (chr)
- `employer`: Contributor employer (chr)
- `occupation`: Contributor occupation (chr)

### 7.1 Data Preparation

```{r}
# Load the dataset
finance <- read_csv("data/campaign_finance_2024.csv")

# Explore the data
glimpse(finance)

```

### 7.2 Creating Contribution Categories

Ask Claude to help you:
- Identify small donors vs major donors (< mean or > mean)
- Create variables for in-state vs out-of-state contributions (assume that your state is New York)

```{r}
# Calculate mean for threshold
mean_amount <- mean(finance$amount, na.rm = TRUE)
# Create donor categories
finance <- finance %>%
mutate(
# Donor size categories based on mean
donor_size = if_else(
amount < mean_amount,
"small_donor",
"major_donor"
),
# In-state vs out-of-state (assuming we're in New York)
contribution_source = if_else(
state == "NY",
"in_state",
"out_of_state"
)
)
# Summarize results
finance %>%
count(donor_size) %>%
mutate(pct = n / sum(n) * 100)

finance %>%
count(contribution_source) %>%
mutate(pct = n / sum(n) * 100)

# Analyze by party
finance %>%
group_by(party, donor_size) %>%
summarise(
total_raised = sum(amount, na.rm = TRUE),
avg_contribution = mean(amount, na.rm = TRUE),
n_contributions = n(),
.groups = "drop"
)


```

### 7.3 Time-Based Transformations

Work with Claude to summarize the data.
- When was the most money given? Contribution timing (early, middle, late in campaign)
- How much money was given each month? 
- How much was contributed over time

```{r}
# Convert date and create time variables
finance <- finance %>%
mutate(
date = as.Date(date),
month = month(date, label = TRUE),
month_num = month(date),
# Create campaign period categories
campaign_period = case_when(
month_num <= 3 ~ "early",
month_num <= 7 ~ "middle",
TRUE ~ "late"
)
)
# When was the most money given?
finance %>%
group_by(campaign_period) %>%
summarise(
total = sum(amount, na.rm = TRUE),
avg = mean(amount, na.rm = TRUE),
n = n()
) %>%
arrange(desc(total))
```

## 8. Demographic Transformations

**Dataset: data/census_political.csv**

**Description**: Census data merged with political outcomes.

**Variables**:
- `county_id`: County FIPS code (int)
- `county_name`: County name (chr)
- `state`: State abbreviation (chr)
- `population`: Total population (int)
- `median_age`: Median age (dbl)
- `pct_college`: Percent with college degree (dbl)
- `median_income`: Median household income (dbl)
- `unemployment_rate`: Unemployment percentage (dbl)
- `dem_vote_share_2020`: Democratic vote share 2020 (dbl)
- `turnout_2020`: Voter turnout 2020 (dbl)

### 8.1 Initial Exploration

```{r}
# Load the dataset
census <- read_csv("data/census_political.csv")

# Look at the data
glimpse(census)
```

### 8.2 Creating Composite Indicators

Work with Claude to:
- Create urban/rural classifications based on population density (above or below the median)
- Create a variable that captures the quartiles of unemployment 

```{r}
# Calculate median population for urban/rural classification
median_pop <- median(census$population, na.rm = TRUE)
# Create classifications
census <- census %>%
mutate(
# Urban/rural classification based on population (proxy for density)
urban_rural = if_else(
population > median_pop,
"urban",
"rural"
),
# Unemployment quartiles
unemployment_quartile = case_when(
unemployment_rate <= quantile(unemployment_rate, 0.25, na.rm = TRUE) ~ "Q1 (lowest)",
unemployment_rate <= quantile(unemployment_rate, 0.50, na.rm = TRUE) ~ "Q2",
unemployment_rate <= quantile(unemployment_rate, 0.75, na.rm = TRUE) ~ "Q3",
TRUE ~ "Q4 (highest)"
)
)
# Check distributions
census %>%
count(urban_rural) %>%
mutate(pct = n / sum(n) * 100)
```

# Segment B — Causality

# Non-AI Exercises

## 1. Three Requirements for Causality

### 1.1 Match: Causal Requirements

Match each requirement with its definition:

**Requirements:**
a) Association
b) Temporal ordering
c) No confounding

**Definitions:**
1. The cause must come before the effect
2. No third variable explains both X and Y
3. X and Y must be statistically related

Matches: a = __3___, b = __1___, c = __2___

### 1.2 Multiple Choice: Ice Cream and Crime

Why is "ice cream sales cause crime" a spurious correlation?

a) Ice cream and crime are not associated
b) Crime happens before ice cream sales
c) Temperature affects both ice cream sales and crime
d) The correlation is too weak

Answer: ___c__

### 1.3 True or False: Causation

Mark each statement as True (T) or False (F):

__F___ Correlation always implies causation
__T___ Causation requires temporal ordering
__F___ Large correlations are more likely to be causal
__T___ Randomized experiments help establish causation
__F___ Observational data can never show causation

### 1.4 Multiple Choice: What Correlation Alone Can Tell Us

We find that counties with more coffee shops have higher voter turnout. Even before establishing whether coffee shops *cause* turnout, this correlation is still useful because it lets us:

a) Conclude that opening more coffee shops would raise turnout
b) Predict turnout — knowing a county's coffee-shop density helps us anticipate its turnout
c) Recommend coffee shops as a get-out-the-vote strategy
d) Rule out any confounding variables

Answer: __B___

## 2. Identifying Confounders

### 2.1 Fill in the Blanks: Confounding

A confounding variable:

1. Affects both the __cause___ and the __effect___
2. Creates a __spurious___ relationship between X and Y
3. Must occur __prior___ to both X and Y
4. Can be controlled through __random___ assignment
5. Makes causal inference _difficult____

Word bank: treatment, outcome, spurious, prior, random, difficult

### Match: Examples

Match each scenario with the likely confounding variable:

**Scenarios:**
a) Wealthy people live longer
b) Students with more books at home do better in school
c) Cities with more police have more crime
d) Married people are happier

**Confounders:**
1. Crime rates drive police hiring
2. Access to healthcare
3. Parents' education and income
4. Selection into marriage

Matches: a = __2___, b = __3___, c = __1___, d = __d___

## 3. Temporal Ordering

### 3.1 Multiple Choice: Time Order

Which scenario violates temporal ordering for causation?

a) Studying causes better grades
b) Rain causes wet streets
c) High crime causes more police hiring last year
d) Tall children have tall parents

Answer: __d___

### 3.2 Spot the Error: Causal Claims

What's wrong with this causal claim?

"States with stricter gun laws have more gun violence, so gun laws cause violence."

Problem: __ Maybe strict gun laws are in response to existing already high violence, temporal order issue___

### 3.3 Spot the Error: Post Hoc Reasoning

A mayor argues: "We adopted a new policing policy in 2021, and crime fell every year afterward — the policy cut crime." The data show crime had been falling by about the same amount each year *before* 2021 as well.

What reasoning error is this, and why does the timing alone not prove the policy worked?

Problem: __post hoc, ergo propter hoc fallacy___

## 4. Alternative Explanations

### 4.1 Multiple Choice: Spurious Correlations

The correlation between divorce rate and margarine consumption is likely:

a) Causal - margarine causes divorce
b) Causal - divorce causes margarine consumption
c) Spurious - both follow similar time trends
d) Reversed - the causation goes backwards

Answer: __c___

### 4.2 True or False: Research Design

Mark each statement as True (T) or False (F):

__F___ Random assignment eliminates all confounders
__F___ Large samples guarantee causal inference
__T___ Natural experiments can establish causation
__T___ Correlation is necessary but not sufficient for causation
__F___ Time series data always shows causation

### 4.3 Match: Causal Mechanisms

Match each political example with its mechanism:

**Examples:**
a) Education → Higher income
b) Campaign spending → Votes
c) Media coverage → Public opinion
d) Voting laws → Turnout

**Mechanisms:**
1. Increased name recognition
2. Skills and credentials
3. Agenda setting and framing
4. Access and convenience

Matches: a = __2__, b = __1___, c = __3___, d = __4___

### 4.4 Multiple Choice: Selecting on the Outcome

A researcher studies **only** completed suicide-terror campaigns and finds that nearly all occurred under a foreign military occupation, concluding that occupation causes suicide terrorism. Why is this reasoning flawed?

a) Foreign occupation and suicide terrorism are not associated
b) By studying only cases where terrorism occurred (selecting on the outcome), the researcher never sees the many occupations that produced *no* terrorism
c) The correlation is too weak to interpret
d) Suicide terrorism causes occupation, not the other way around

Answer: __b___

# AI Exercises

**Tips for Working with Claude:**

- Ask for **R code using only tidyverse** (no other packages)
- Request **simple, focused answers** to your specific question—not complex analyses
- Ask Claude to **explain what the code is doing** since you're learning
- Avoid asking for visualizations or plots in these exercises
- Include the output of `glimpse()` in your prompt so Claude knows your variable names

**Example prompt:** "Using tidyverse in R, calculate the correlation between min_wage and unemployment_rate to test for association. Keep the code simple and explain what each line does. Here is what my data looks like: [paste glimpse output]"

For each AI exercise:
- Work with Claude to analyze the data
- Record your prompts and key findings

## 5. Testing Causal Claims

**Dataset: data/minimum_wage_effects.csv**

**Description**: State-level data on minimum wage changes and employment.

**Variables**:

- `state`: State name (chr)
- `year`: Year of observation (int)
- `min_wage`: State minimum wage (dbl)
- `unemployment_rate`: Unemployment percentage (dbl)
- `teen_employment_rate`: Teen employment percentage (dbl)
- `median_income`: State median income (dbl)
- `cost_of_living`: Cost of living index (dbl)
- `gdp_growth`: State GDP growth rate (dbl)
- `border_state_wage`: Average minimum wage in bordering states (dbl)

### 5.1 Initial Exploration

```{r}
# Load the dataset
wage_data <- read_csv("data/minimum_wage_effects.csv")

# Examine the structure
glimpse(wage_data)
```

### 5.2 Checking Causal Requirements

Ask Claude to help you:
- Test for association between minimum wage and employment

```{r}
cor.test(wage_data$min_wage, wage_data$unemployment_rate)
```

### 5.3 Alternative Explanations

Work with Claude to explore alternative explanations for any patterns you find.

## 6. Media Effects on Politics

**Dataset: data/media_influence_study.csv**

**Description**: Data on media consumption and political attitudes.

**Variables**:

- `respondent_id`: Unique identifier (int)
- `survey_wave`: Wave 1 or 2 (int)
- `fox_news_hours`: Weekly hours watching Fox News (dbl)
- `msnbc_hours`: Weekly hours watching MSNBC (dbl)
- `conservative_score`: Conservative attitudes, 0-100 (dbl)
- `trump_approval`: Trump approval, 0-100 (dbl)
- `age`: Age in years (int)
- `education`: Education level (chr)
- `party_id`: Party identification (chr)
- `political_interest`: Interest in politics, 1-10 (int)

### 6.1 Media and Attitudes

```{r}
# Load the dataset
media_data <- read_csv("data/media_influence_study.csv")

# Check the structure
glimpse(media_data)
```

### 6.2 Reverse Causation

Work with Claude to:
- Test whether conservatives choose Fox News (selection)

```{r}
media_data %>%
filter(survey_wave == 1) %>%
group_by(party_id) %>%
summarize(
mean_fox_hours = mean(fox_news_hours),
mean_msnbc_hours = mean(msnbc_hours)
)
```

### 6.3 Confounding Factors

Ask Claude to help identify and control for variables that might affect both media choice and political attitudes.

## 7. Campaign Finance and Elections

**Dataset: data/campaign_causality.csv**

**Description**: Congressional election data with spending information.

**Variables**:

- `district_id`: Congressional district (chr)
- `year`: Election year (int)
- `incumbent_party`: Party of incumbent (chr)
- `incumbent_spending`: Incumbent campaign spending (dbl)
- `challenger_spending`: Challenger campaign spending (dbl)
- `incumbent_vote_share`: Incumbent's vote percentage (dbl)
- `presidential_vote`: Presidential vote in district (dbl)
- `median_income`: District median income (dbl)
- `incumbent_scandal`: Whether incumbent had scandal (lgl)
- `quality_challenger`: Whether challenger held prior office (lgl)

### 7.1 Money and Votes

```{r}
# Load the dataset
campaign_data <- read_csv("data/campaign_causality.csv")

# Explore the data
glimpse(campaign_data)
```

### 7.2 Endogeneity Problems

Ask Claude to help you understand:
- Why might successful candidates raise more money?
- How does expected competitiveness affect fundraising?
- What confounders affect both money and votes?

### 7.3 Natural Experiments

Work with Claude to identify situations where spending might be more randomly assigned (e.g., unexpected donations, spending limits).

## 8. Policy Evaluation

**Dataset: data/policy_intervention.csv**

**Description**: Data on states before/after policy changes.

**Variables**:

- `state`: State name (chr)
- `year`: Year (int)
- `policy_adopted`: Whether state adopted policy (lgl)
- `year_adopted`: Year policy was adopted (int)
- `outcome_variable`: Policy outcome of interest (dbl)
- `population`: State population (int)
- `unemployment`: Unemployment rate (dbl)
- `gov_party`: Governor's party (chr)
- `leg_control`: Party controlling legislature (chr)
- `neighbor_adopted`: Whether neighboring state has policy (lgl)

### 8.1 Policy Effects

```{r}
# Load the dataset
policy_data <- read_csv("data/policy_intervention.csv")

# Look at the structure
glimpse(policy_data)
```

### 8.2 Selection into Treatment

Work with Claude to:
- Identify which states adopted the policy and when
- Check pre-treatment trends

```{r}
policy_data %>%
group_by(policy_adopted) %>%
summarize(
mean_outcome = mean(outcome_variable),
mean_unemployment = mean(unemployment),
n = n()
)
```

\newpage
