---
title: "Week 5 · Class 1: Practice Exercises"
subtitle: "Modern Causal Inference & Linear Regression"
date: "2026"
format:
  pdf:
    toc: false
    number-sections: true
    colorlinks: true
---

```{r}
#| include: false
library(tidyverse)
library(knitr)
library(formatR)
# When running chunks interactively, set the working directory to this file's
# folder so the relative data/ paths resolve. (During render, knitr already
# uses the document's directory, so this is skipped.)
if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
  .doc_path <- rstudioapi::getSourceEditorContext()$path
  if (length(.doc_path) == 1 && nzchar(.doc_path)) setwd(dirname(.doc_path))
}
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = FALSE)
```

# Segment A — Modern Causal Inference

# Non-AI Exercises

## 1. Fundamental Problem of Causal Inference

### 1.1 Multiple Choice: Counterfactuals

What is the fundamental problem of causal inference?

a) We can't measure outcomes accurately
b) We can't observe the same unit under both treatment and control conditions
c) Random assignment is difficult
d) Statistical software has limitations

Answer: _____

### 1.2 Fill in the Blanks: Potential Outcomes

In potential outcomes notation:

1. Y(1) represents the outcome when _____
2. Y(0) represents the outcome when _____
3. The individual causal effect is _____
4. We can only observe _____ of these for each person
5. The unobserved outcome is called a _____

Word bank: treated, not treated, Y(1) - Y(0), one, counterfactual

### 1.3 True or False: Causal Inference

Mark each statement as True (T) or False (F):

_____ We can directly observe individual causal effects
_____ Random assignment helps create comparable groups
_____ The ATE is the average of individual treatment effects
_____ Counterfactuals are hypothetical unobserved outcomes
_____ Causal inference requires observing all potential outcomes

## 2. Average Treatment Effects

### 2.1 Code Detective: ATE Calculation

What does this code calculate?

```r
treatment_group %>%
  summarise(mean_outcome = mean(outcome)) -> treated_mean

control_group %>%
  summarise(mean_outcome = mean(outcome)) -> control_mean

ate <- treated_mean$mean_outcome - control_mean$mean_outcome
```

This calculates: _____

### 2.2 Match: ATE Components

Match each component with its meaning:

**Components:**
a) E[Y(1)]
b) E[Y(0)]
c) E[Y(1) - Y(0)]
d) E[Y|T=1] - E[Y|T=0]

**Meanings:**
1. Average treatment effect
2. Expected outcome under treatment
3. Expected outcome under control
4. Observed difference in means

Matches: a = _____, b = _____, c = _____, d = _____

### 2.3 Multiple Choice: Random Assignment

Why does random assignment help identify causal effects?

a) It makes the sample larger
b) It ensures treatment and control groups are similar on average
c) It eliminates all confounders perfectly
d) It makes people more honest

Answer: _____

## 3. Difference-in-Differences

### 3.1 Fill in the Blanks: DiD Logic

Difference-in-differences compares:

1. Changes over time in the _____ group
2. Changes over time in the _____ group
3. The difference between these changes estimates the _____
4. DiD assumes _____ trends without treatment
5. We need at least _____ time periods

Word bank: treatment, control, causal effect, parallel, two

### 3.2  True or False: DiD Assumptions

Mark each statement as True (T) or False (F):

_____ DiD requires random assignment of treatment
_____ Parallel trends assumption means groups move together pre-treatment
_____ DiD can work with observational data
_____ We need panel data for DiD
_____ DiD eliminates all time-invariant confounders

## 4. Validity

### 4.1 Match: Validity Types

Match each validity type with its concern:

**Types:**
a) Internal validity
b) External validity
c) Construct validity
d) Statistical validity

**Concerns:**
1. Whether results generalize to other contexts
2. Whether we can attribute effects to treatment
3. Whether we're measuring what we think
4. Whether statistical tests are appropriate

Matches: a = _____, b = _____, c = _____, d = _____

### 4.2 Multiple Choice: Threats

Which threatens internal validity?

a) Small sample size
b) Selection bias
c) Limited geographic scope
d) Measurement error

Answer: _____

### 4.3 Spot the Error: Causal Claims

What's wrong with this conclusion?

"States that adopted the policy saw crime decrease, so the policy reduced crime."

Problem: _____

# AI Exercises

**Tips for Working with Claude:**

- Ask for **R code using only tidyverse** (no other packages)
- Request **simple, focused answers** to your specific question—not complex analyses
- Ask Claude to **explain what the code is doing** since you're learning
- Avoid asking for visualizations or plots in these exercises
- Include the output of `glimpse()` in your prompt so Claude knows your variable names

**Example prompt:** "Using tidyverse in R, calculate the average treatment effect (ATE) by comparing mean post_earnings_12m between treated and control groups. Keep the code simple and explain what each line does. Here is what my data looks like: [paste glimpse output]"

For each dataset below, work with Claude to:
- Calculate the Average Treatment Effect (ATE)
- Test for statistical significance
- Interpret the magnitude and meaning of the ATE in context

## 5. Job Training Program RCT

**Dataset: data/job_training_rct.csv**

**Description**: Randomized controlled trial of a job training program with 1,200 participants.

**Variables**:

- `participant_id`: Unique identifier (int)
- `age`: Age in years (int)
- `gender`: Gender (chr)
- `education`: Education level 1-5 (int)
- `prior_earnings`: Annual earnings before program (dbl)
- `treated`: Treatment assignment 1=treated, 0=control (int)
- `takeup`: Actually participated if treated (int)
- `post_earnings_6m`: Earnings 6 months after (dbl)
- `post_earnings_12m`: Earnings 12 months after (dbl)
- `employment_12m`: Employed at 12 months 1=yes (int)
- `center_id`: Training center ID (int)

```{r}
# Load the data
job_data <- read_csv("data/job_training_rct.csv")

# Calculate ATE for earnings at 12 months

# Calculate ATE for employment at 12 months
```

Interpret: What do these ATEs mean in practical terms?

## 6. Get Out The Vote Experiment

**Dataset: data/gotv_experiment.csv**

**Description**: Field experiment testing different voter mobilization methods with 5,000 registered voters.

**Variables**:

- `voter_id`: Unique identifier (int)
- `treatment`: Treatment group - "Control", "Phone Call", "Canvassing", "Mailer" (chr)
- `age_group`: Age category (chr)
- `education`: Education level (chr)
- `voted_2022`: Whether voted in 2022 election 1=yes (int)

```{r}
# Load the data
gotv_data <- read_csv("data/gotv_experiment.csv")

glimpse(gotv_data)

# Calculate ATE for each treatment compared to control

```

Interpret: Which intervention is most effective? Is it cost-effective?

## 7. Media Advertising Experiment

**Dataset: data/media_experiment_full.csv**

**Description**: Experiment on political advertising effects with 1,400 participants tracked over time.

**Variables**:

- `viewer_id`: Unique identifier (int)
- `day`: Day of observation (int)
- `battleground_state`: In battleground state 1=yes (int)
- `minutes_tv_news`: Minutes of TV news watched (int)
- `total_ads_seen`: Total political ads seen (int)
- `voting_intent`: Voting intention 0-100 scale (dbl)
- `ads_candidate_A`: Number of ads for candidate A (int)
- `ads_candidate_B`: Number of ads for candidate B (int)

```{r}
# Load the data
media_data <- read_csv("data/media_experiment_full.csv")

# Calculate ATE of seeing more ads for candidate A vs B on voting intent

```

Interpret: How much do political ads move voter preferences?

## 8. Online Course Completion

**Dataset: data/mooc_progress.csv**

**Description**: Data from 3,000 students in online courses comparing free vs paid subscriptions.

**Variables**:

- `enroll_id`: Enrollment ID (int)
- `day_of_course`: Day in course (int)
- `videos_watched`: Number of videos watched (int)
- `quiz_attempted`: Attempted quiz 1=yes (int)
- `forum_posts`: Number of forum posts (int)
- `subscription_type`: "free" or "paid" (chr)
- `completed_course`: Completed course 1=yes (int)

```{r}
# Load the data
mooc_data <- read_csv("data/mooc_progress.csv")

glimpse(mooc_data)

# Calculate ATE of paid subscription on course completion

```

Does paying for a course increase commitment to complete it?

## 9. Refugee Integration Program

**Dataset: data/refugee_integration.csv**

**Description**: Study of 800 refugee families comparing different levels of English language training.

**Variables**:

- `case_id`: Family case ID (int)
- `arrival_year`: Year arrived (int)
- `country_origin`: Country of origin (chr)
- `family_size`: Number of family members (int)
- `english_class_hours`: Hours of English classes provided (int)
- `employed_year1`: Family member employed by year 1 (int)
- `income_year5`: Family income at year 5 (dbl)
- `citizenship_applied`: Applied for citizenship 1=yes (int)

```{r}
# Load the data
refugee_data <- read_csv("data/refugee_integration.csv")

glimpse(refugee_data)
# Calculate ATE of receiving >100 hours vs ≤100 hours of English training on employment

```

Is intensive English training worth the investment?

# Segment B — Linear Regression

# Non-AI Exercises

## 1. Understanding Linear Relationships

### 1.1 Multiple Choice: Linear Relationships

What defines a linear relationship between X and Y?

a) X and Y move in the same direction
b) As X increases by one unit, Y changes by a constant amount
c) The relationship forms a perfect straight line
d) X causes Y

Answer: _____

### 1.2 True or False: Regression Concepts

Mark each statement as True (T) or False (F):

_____ Linear regression finds the line that minimizes squared errors
_____ The slope tells us the change in Y for a one-unit change in X
_____ The intercept is the predicted Y when X equals zero
_____ R-squared can be negative
_____ Residuals are the differences between predicted and actual values

### 1.3 Match: Regression Components

Match each component with its symbol:

**Components:**

a) Dependent variable
b) Independent variable
c) Slope coefficient
d) Intercept
e) Error term

**Symbols:**

1. $\beta$
2. Y
3. $\mu$
4. X
5. $\alpha$

Matches: a = _____, b = _____, c = _____, d = _____, e = _____

## 2. Interpreting Regression Output

### 2.1 Code Detective: lm() Function

What does this code do?

```r
model <- lm(vote_share ~ campaign_spending, data = election_data)
summary(model)
```

This code: 

### 2.2 Fill in the Blanks: Interpretation

Given a regression equation: Y = 45 + 3.2X

1. The intercept is _____
2. The slope is _____
3. When X = 0, Y is predicted to be _____
4. When X increases by 1, Y increases by _____
5. When X = 10, Y is predicted to be _____

## 3. Assumptions and Diagnostics

### 3.1 Match: OLS Assumptions

Match each assumption with its meaning (you will need to look each up!):

**Assumptions:**

a) Linearity
b) Independence
c) Homoscedasticity
d) Normality

**Meanings:**

1. Residuals have constant variance
2. The true relationship is linear
3. Observations are independent
4. Residuals are normally distributed

Matches: a = _____, b = _____, c = _____, d = _____

## 4. Model Evaluation

### 4.1 Multiple Choice: RMSE

Root Mean Square Error (RMSE) measures:

a) The average error in predictions
b) The typical size of prediction errors
c) The percentage of variance explained
d) The significance of coefficients

Answer: _____

### 4.2 True or False: Model Quality

Mark each statement as True (T) or False (F):

_____ Lower RMSE indicates better predictions
_____ Outliers can strongly affect regression results
_____ The best-fit line always goes through (mean(X), mean(Y))

# AI Exercises

**Tips for Working with Claude:**

- Ask for **R code using only tidyverse** (no other packages)
- Request **simple, focused answers** to your specific question—not complex analyses
- Ask Claude to **explain what the code is doing** since you're learning
- Avoid asking for visualizations or plots in these exercises
- Include the output of `glimpse()` in your prompt so Claude knows your variable names

**Example prompt:** "Using tidyverse in R, run a simple linear regression with lm() predicting approval from unemployment. Then show me how to use the coefficients to predict approval when unemployment is 8%. Keep the code simple and explain what each line does. Here is what my data looks like: [paste glimpse output]"

For each AI exercise:
- Work with Claude to analyze the data
- Record your prompts and key findings

## 5. Political Approval and Economy

**Dataset: data/approval_economy.csv**

**Description**: Presidential approval ratings and economic indicators.

**Variables**:

- `month`: Month of observation (date)
- `approval`: Presidential approval percentage (dbl)
- `unemployment`: Unemployment rate (dbl)
- `gdp_growth`: GDP growth rate (dbl)
- `inflation`: Inflation rate (dbl)
- `gas_prices`: Average gas price (dbl)
- `consumer_confidence`: Consumer confidence index (dbl)
- `party`: President's party (chr)
- `major_event`: Major event that month (chr)

### 5.1 Initial Regression Analysis

```{r}
# Load the dataset
approval_data <- read_csv("data/approval_economy.csv")

# Examine the structure
glimpse(approval_data)
```

### 5.2 Work with Claude to run a simple regression of approval on unemployment.

```{r}

```

### 5.3 Use your regression model to predict presidential approval when unemployment is 8%.

```{r}

```

## 6. Campaign Finance Effects

**Dataset: data/campaign_regression.csv**

**Description**: Congressional election results with spending data.

**Variables**:

- `district`: Congressional district ID (chr)
- `incumbent_vote`: Incumbent vote share percentage (dbl)
- `incumbent_spend`: Incumbent spending in millions (dbl)
- `challenger_spend`: Challenger spending in millions (dbl)
- `total_spend`: Total spending in race (dbl)
- `incumbent_party`: Party of incumbent (chr)
- `presidential_margin`: Presidential vote margin in district (dbl)
- `incumbent_tenure`: Years in office (int)
- `primary_challenge`: Had primary challenger (lgl)

### 6.1 Money and Votes

```{r}
# Load the dataset
campaign_data <- read_csv("data/campaign_regression.csv")

# Check the structure
glimpse(campaign_data)
```

### 6.2 Use Claude to run a simple regression of incumbent vote share on incumbent spending.

```{r}

```

### 6.3 Use your regression model to predict incumbent vote share when the incumbent spends $7 million.

```{r}

```

## 7. Education and Political Knowledge

**Dataset: data/education_knowledge.csv**

**Description**: Survey data on education and political knowledge.

**Variables**:

- `respondent_id`: Unique identifier (int)
- `years_education`: Years of formal education (int)
- `political_knowledge`: Score on political knowledge test, 0-100 (dbl)
- `age`: Age in years (int)
- `income`: Annual income in thousands (dbl)
- `news_consumption`: Hours of news per week (dbl)
- `college_major`: Field of study if college graduate (chr)
- `parents_education`: Average years of parents' education (dbl)
- `books_at_home`: Number of books growing up (int)

### 7.1 Education Effects

```{r}
# Load the dataset
education_data <- read_csv("data/education_knowledge.csv")

# Explore the data
glimpse(education_data)
```

### 7.2 Work with Claude to estimate the relationship between education and political knowledge.

```{r}

```

### 7.3 Use your regression model to predict political knowledge for someone with 16 years of education (college degree).

```{r}

```

## 8. Media Bias and Perception

**Dataset: data/media_perception.csv**

**Description**: Data on media consumption and political perceptions.

**Variables**:

- `respondent_id`: Unique identifier (int)
- `fox_hours`: Weekly hours watching Fox News (dbl)
- `msnbc_hours`: Weekly hours watching MSNBC (dbl)
- `newspaper_freq`: Days per week reading newspaper (int)
- `social_media_news`: Hours on social media for news (dbl)
- `perceived_bias`: Perceived media bias, -10 (left) to 10 (right) (dbl)
- `actual_knowledge`: Political knowledge score (dbl)
- `confidence`: Confidence in political knowledge, 0-100 (dbl)
- `party_id`: Party identification (chr)

### 8.1 Media Diet Analysis

```{r}
# Load the dataset
media_data <- read_csv("data/media_perception.csv")

# Look at the structure
glimpse(media_data)
```

### 8.2 Use Claude to run a simple regression of perceived bias on Fox News hours watched.

```{r}

```

### 8.3 Use your regression model to predict perceived bias for someone who watches 10 hours of Fox News per week.

```{r}

```

\newpage
