---
title: "Week 6 · Class 1: Practice Exercises"
subtitle: "Plotting Data"
date: "2026"
format:
  pdf:
    toc: false
    number-sections: true
    colorlinks: true
---

```{r}
#| include: false
library(tidyverse)
library(knitr)
library(formatR)
# When running chunks interactively, set the working directory to this file's
# folder so the relative data/ paths resolve. (During render, knitr already
# uses the document's directory, so this is skipped.)
if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
  .doc_path <- rstudioapi::getSourceEditorContext()$path
  if (length(.doc_path) == 1 && nzchar(.doc_path)) setwd(dirname(.doc_path))
}
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = FALSE)
```

# Non-AI Exercises

## 1. Visualization Principles

### 1.1 Multiple Choice: Chart Types

Which visualization is best for showing the relationship between two continuous variables?

a) Bar chart
b) Pie chart
c) Scatter plot
d) Box plot

Answer: _____

### 1.2 Match: Plot Types and Uses

Match each plot type with its best use:

**Plot Types:**

a) Histogram
b) Line chart
c) Bar chart
d) Box plot
e) Heatmap

**Uses:**

1. Compare categories
2. Show distribution of one variable
3. Display trends over time
4. Show distributions by group
5. Display correlation matrix

Matches: a = _____, b = _____, c = _____, d = _____, e = _____

### 1.3 True or False: Visualization Best Practices

Mark each statement as True (T) or False (F):

_____ Always start y-axis at zero for bar charts  
_____ Pie charts are ideal for comparing many categories  
_____ Color should be used meaningfully, not decoratively  
_____ 3D effects improve data comprehension  
_____ Labels should be clear and descriptive

## 2. Grammar of Graphics

### 2.1 Fill in the Blanks: ggplot Components

The main components of a ggplot are:

1. _____ : The dataset  
2. _____ : Variables mapped to visual properties  
3. _____ : The type of plot (points, lines, bars)  
4. _____ : Axis labels and title  
5. _____ : Overall visual style

Word bank: data, aesthetics (aes), geometries (geom), labels, theme

### 2.2 What does each line add to the plot?

```r
ggplot(data, aes(x = income, y = vote_share)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~region) +
  theme_minimal()
```

Line 2: _____  
Line 3: _____  
Line 4: _____  
Line 5: _____

### 2.3 Multiple Choice: Aesthetics

In ggplot, aes() can map variables to:

a) Colors only
b) X and Y positions only
c) Multiple visual properties (color, size, shape, etc.)
d) Statistical transformations

Answer: _____

## 3. Effective Communication

### 3.1 Spot the Error: Bad Visualization

What's wrong with using a pie chart to show changes over 50 time periods?

Problem: _____

### 3.2 Match: Color Choices

Match color use with purpose:

**Color Uses:**

a) Sequential palette  
b) Diverging palette  
c) Categorical palette  
d) Highlight color

**Purposes:**

1. Show distinct unordered groups  
2. Emphasize one category  
3. Show values from low to high  
4. Show values around a midpoint

Matches: a = _____, b = _____, c = _____, d = _____

## 4. Advanced Visualization

### 4.1 True or False: Faceting

About faceting in ggplot:

_____ facet_wrap() creates a grid of plots  
_____ facet_grid() requires two variables  
_____ Facets must share the same scale  
_____ You can combine faceting with grouping  
_____ Faceting helps avoid overplotting  

### 4.2 Multiple Choice: Handling Overplotting

When you have too many points overlapping, you should:

a) Make points smaller  
b) Use transparency (alpha)  
c) Use jittering or binning  
d) All of the above

Answer: _____

# AI Exercises

**Tips for Working with Claude:**

- Ask for **R code using only tidyverse and ggplot2** (no other packages)
- Request **simple, focused answers** to your specific question—not complex analyses
- Ask Claude to **explain what the code is doing** since you're learning
- For this class, you CAN ask for visualizations—that's the focus!
- Include the output of `glimpse()` in your prompt so Claude knows your variable names

**Example prompt:** "Using ggplot2 in R, create a line chart showing dem_vote_share over time (year) with different colors for each region. Keep the code simple and explain what each line does. Here is what my data looks like: [paste glimpse output]"

For each AI exercise:

- Work with Claude to create visualizations
- Record your prompts and describe the plots created

## 5. Election Results Visualization

**Dataset: data/election_visualization.csv**

**Description**: State-level election results over time.

**Variables**:

- `state`: State name (chr)
- `year`: Election year (int)
- `dem_vote_share`: Democratic vote percentage (dbl)
- `rep_vote_share`: Republican vote percentage (dbl)
- `turnout`: Voter turnout percentage (dbl)
- `swing_state`: Swing state indicator (lgl)
- `region`: Geographic region (chr)
- `electoral_votes`: Number of electoral votes (int)
- `margin`: Victory margin (dbl)
- `incumbent_party`: Party of incumbent (chr)

### 5.1 Time Series Visualization

```{r}
# Load the dataset
election_data <- read_csv("data/election_visualization.csv")

# Examine the structure
glimpse(election_data)
```

### 5.2 Geographic Patterns

Work with Claude to create time series plots showing partisan trends over time

Ask Claude to help you:
- Create regional comparisons
- Visualize swing states differently
- Show electoral vote weights

```{r}

```

## 6. Survey Response Patterns

**Dataset: data/survey_visualization.csv**

**Description**: Large political survey with demographic breakdowns.

**Variables**:

- `respondent_id`: Unique identifier (int)
- `age_group`: Age category (chr)
- `education`: Education level (chr)
- `income_bracket`: Income category (chr)
- `party_id`: Party identification (chr)
- `ideology`: Liberal-conservative scale 1-7 (int)
- `approval_president`: Presidential approval 0-100 (dbl)
- `approval_congress`: Congressional approval 0-100 (dbl)
- `trust_media`: Media trust 0-100 (dbl)
- `political_interest`: Interest level 1-10 (int)

### 6.1 Distribution Visualizations

```{r}
# Load the dataset
survey_data <- read_csv("data/survey_visualization.csv")

# Check the structure
glimpse(survey_data)
```

### 6.2 Group Comparisons

Use Claude to create effective distribution plots for different variables of your choice.

```{r}

```

## 7. Legislative Behavior

**Dataset: data/legislative_visualization.csv**

**Description**: Congressional voting patterns and member characteristics.

**Variables**:

- `member_id`: Legislator identifier (int)
- `name`: Member name (chr)
- `party`: Political party (chr)
- `state`: State abbreviation (chr)
- `chamber`: House or Senate (chr)
- `ideology_score`: DW-NOMINATE score (dbl)
- `votes_with_party`: Percentage voting with party (dbl)
- `bills_sponsored`: Number of bills sponsored (int)
- `committee_leadership`: Leadership positions (int)
- `tenure`: Years in office (int)

### 7.1 Ideology Visualization

```{r}
# Load the dataset
legislative_data <- read_csv("data/legislative_visualization.csv")

# Explore the data
glimpse(legislative_data)
```

### 7.2 Bivariate Relationships

Work with Claude to visualize ideological distributions and polarization.

```{r}

```
